\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-12-06 latex/paper.tex
chore: redo plot and new intermediate results benchmark

% author info
\author{Lionel J. Garcia}
\author{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}
\author{Dax L. Feliz}
\author{Catriona A. Murray}
\correspondingauthor{Lionel J. Garcia}
\email{lgarcia@flatironinstitute.org}

@@

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with them, a common approach consists in cleaning a light curve from nuisance signals before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signal-to-noise-ratio down to no detection. As this degradation depends on the correlated noise characteristics, we explore the relative parameter space for which transits are altered, and quantify this effect on a set of simulated light curves. On this synthetic dataset, we show that \nuancemethod{} largely outperforms the detection capabilities of a commonly used transit search technique, which consists in cleaning light curves from styellar variability with an optimal bi-weight filter before searching for transits using a box-least-square algorithm. Then, in order to assess the performance of \nuancemethod{} on real datasets, we inject and recover transits into the TESS light curves of 438 M dwarfs, and compare our method to 4 detrending techniques followed by a Box-Least-Square search. For transits with a duration exceeding one fifth of the stellar variability timescale, i.e. when both signals start resembling each other, we show that our algorithm is the most performant in 94\% of cases, leading to both the highest number of true positives and the lowest number of false positive detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationaly expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices (?). Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of active stars and their transiting exoplanets, both using space-based and sparse ground-based observations.
\end{abstract}

@@

As seen in \autoref{fig:toi540_periodograms}, the SNR periodogram using \texttt{biweight+BLS} and \texttt{nuance} are very similar, with the known transiting exoplanet TOI-540 b detected with an orbital period P = 1.24 days. This is well expected as the relative parameter $\tau$ equal 13 for TOI-540 b \footnote{using \autoref{eq:relative_params} with the stellar rotation $P=0.72$ days and the known transit duration $D=29.5$ min of TOI-540 b \citep{TOI540}}, which lies outside the range where \nuancemethod{} is expected to be beneficial beneficial (see \autoref{fig:snr_detrend}). Nonetheless, the \nuance{} periodogram of TOI 540 features less spurious SNR peaks, largely due to the penalty naturally occurring when single transits with different depths are periodically combined. In the second search (right panel) of \autoref{fig:toi540_periodograms}, we also notice a higher number of peaks that would lead to false positive detections of transits in the \texttt{biweight+BLS} case. The proper treatment of correlated noise in \nuance{}, as observed in \autoref{real}, makes these peaks non-significant, avoiding a large number of false detections.\\\\

@@

\subsection{The choice of kernel and model}
While not being discussed in our study, \textbf{the efficiency of \nuancemethod{} to detect transits in correlated noise is highly dependant on the design of its GP kernel}. In the ensemble comparison of \autoref{real}, the goal was to choose a kernel and an optimization strategy suited to most of the studied light curves, leading to few outliers in the results that were indicative of a badly designed and/or optimized kernel. An alternative, recommended for more realistic blind searches, is to perform model comparison on well selected kernels, and to adapt the optimization strategy to each dataset.\\\\
When using \nuancecode{} on TESS light curves for example, it must be noted that the observed light curve variability might originate from a contaminated photometric aperture, so that a physically-interpretable GP kernel representing a single star activity is not necessarily appropriate. On the other hand, a single squared exponential GP kernel might also be sufficient for some applications.\\\\
Something to emphasize is that \nuancemethod{} cannot be used to produce reliable detrended light curves. Most of the time, the GP is flexible enough to partially model transits and can never be used to model the data without them. The idea behind \nuancemethod{} is to compute the likelihood of data against a model containing both transits and correlated noise, without ever trying to disentangle both signals. In practice, it means that it can be very hard to actually verify the presence of transits found by \nuancemethod{} visually, so that transits may be detected but remain hidden in correlated noise. This is particularly true for stars displaying very high-frequency photometric pulsations.

\subsection{Prospects}
The present implementation of \nuancemethod{} has the potential to be extended to be used beyond the search of periodic transits. Here are ideas of possible use and extensions, from the most straightforward to the most ambitious:
\begin{enumerate}
    \item In order to compare \nuancecode{} to BLS-based methods, we injected and retrieved only box-shaped transits. However, similarly to the Transit-Least-Square algorithm from \cite{tls}, \textbf{limb-darkened transits can serve as base model} in the linear search and are expected to improve the transit search in the same way TLS provided an improvement over BLS.
    \item The linear search of \nuancemethod{} is a single-event detection algorithm that can be used to search for single transit events, but also \textbf{detect transiting exocomets and flares}, by simply adapting the base astrophysical model in the last colum of the design matrix $\bf{X}$. While not being tested and benchmarked, the \nuancecode{} implementation already integrates this feature and we show an example of a search of the known exocomets of Beta Pictoris in \autoref{fig:exocomets}.
    \item During the periodic search, no prior about the transit duration related to the orbital period of the planet was used. This was done in order to allow the detection of grazing transiting exoplanets that would produce shorter timescale transits compared to what is expected from a circular orbit with a null impact parameter. However, adding such priors might produce less spurious periodogram peaks and be very beneficial for the automatic search of transits in large datasets. Another idea, similar to the one employed in \cite{foreman2016}, is to leverage \textbf{model comparison in order to reject single transits that are better described by the GP model alone}. Both ideas come at no cost given our modeling approach.
    \item Finally, the formalism of \nuancemethod{} could be adapted and used to \textbf{search for exoplanets featuring transit time variations (TTV)}. Indeed, this application only requires a modification of the periodic search, as maximum likelihood peaks close to linearly predicted transit epochs may be considered. This could be done either with a special nearest-neighbor algorithm or with a convolution of the computed likelihood grid with a Gaussian kernel. To maximize efficiency and interpretability, we would recommend these approaches to be explored analytically, rather than using a data-driven treatment of the linear search products.
\end{enumerate}

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{exocomet.pdf}
        \caption{Demonstration of the search of transiting exocomets in the TESS light curves of the star $\beta$ Pictoris (2 minutes PDCSAP data from sector 33). This star is known to display rapid $\delta$ Scuti type photometric variations with a period of about 30 minutes \citep{Lecavelier2022}. Here, we simply used the linear search of \nuancecode{} with a different base model, one that mimics the shape of transiting exocomets, to compute the SNR time-series of the signal over one sector. The maximum-SNR events are displayed in the bottom of the figure and match with the ones found by \cite{Lecavelier2022}. The GP kernel choice and its hyperparameters optimization are similar to the one presented in \autoref{real}.}
        \end{centering}
        \label{fig:exocomets} 
\end{figure}

\section{Conclusion}
This paper presents \nuancemethod{}, an algorithm designed to detect planetary transits in light curves featuring correlated noise in the form of instrumental signals and stellar variability. In this context, a conventional approach involves detrending a light curve before searching for transits using a Box-Least-Square algorithm. However, we show that the incomplete modeling of a light curve for its detrending, one that ignores the presence of transits, degrades their signal-to-noise-ratio down to the point of not being detectable. Adopting a widely used detrending stratgey, we explore the extend of this degradation on simulated light curves, and its dependance on the stellar variability characteristics, showing the need for a full-fledged transit search method like \nuancemethod{}.\\\\
The effectiveness of \nuancemethod{} is tested using a synthetic dataset and further validated on real TESS light curves of 438 rapidly-rotating M dwarfs. These injection-recovery tests reveal that \nuancemethod{} consistently outperforms conventional transit search techniques, especially for transits with duration exceeding one-fifth of the stellar variability timescale. In these cases, \nuancemethod{} not only identifies a higher number of true positive detections but also minimizes false positives, demonstrating its robustness and reliability.\\\\
We make \nuancemethod{} publicly available through the \nuancecode{} open-source Python package. This package leverages the power of JAX, allowing for efficient computations and compatibility with distributed computing environments and GPU devices. Nonetheless we acknowledge the limitations of \nuancemethod{} and its increased computational cost compared to more tr. In this regard, \nuancemethod{} should be used as an alternative to more traditional techniques only in the presence of substantial correlated noise. As guidelines for choosing \nuancemethod{} over other techniques are highly dependant on the type of detrending algorithm employed, we reserve this study to a future work.\\\\
Finally, we suggest future improvements and extensions of the algorithm, including its application for detecting single transits, exocomets, flares, and transit time variations (TTV), underscoring its versatility and potential for broader impact in astronomical research.
\\\\
\vfill{}
We would like to thank Julien De Wit, Prajwal Niraula and Francisco J. Pozuelos for meaningful discussions at the beginning of this project, Germain Garcia for his useful insights about numerical optimization, Michaël Gillon for his overall support, and the member of the Astronomical Data Group at the Center for Computational Astrophysics for many enriching discussions and feedback.x

>>>>

% author info
\author{Lionel J. Garcia}
\author{Daniel Foreman-Mackey}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY, USA}
\author{Dax L. Feliz}
\author{Catriona A. Murray}
\author{Francisco J. Pozuelos}
\correspondingauthor{Lionel J. Garcia}
\email{lgarcia@flatironinstitute.org}

@@

\begin{abstract}
    We present \nuancemethod{}, an algorithm to search for planetary transits in light curves featuring correlated noise, such as instrumental signals and stellar photometric variability. To deal with them, a common approach consists in cleaning a light curve from nuisance signals before searching for transits. However, we show that this approach, based on the prior assumption that transits are not present, strongly degrades their signal-to-noise-ratio down to no detection. As this degradation depends on the correlated noise characteristics, we explore the relative parameter space for which transits are altered, and quantify this effect on a set of simulated light curves. On this synthetic dataset, we show that \nuancemethod{} largely outperforms the detection capabilities of a commonly used transit search technique, which consists in cleaning light curves from styellar variability with an optimal bi-weight filter before searching for transits using a box-least-square algorithm. Then, in order to assess the performance of \nuancemethod{} on real datasets, we inject and recover transits into the TESS light curves of 438 M dwarfs, and compare our method to 4 detrending techniques followed by a Box-Least-Square search. For transits with a duration exceeding one fifth of the stellar variability timescale, i.e. when both signals start resembling each other, we show that our algorithm is the most performant in 94\% of cases, leading to both the highest number of true positives and the lowest number of false positive detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationaly expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices. Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of active stars and their transiting exoplanets, both using space-based and sparse ground-based observations.
\end{abstract}

@@

As seen in \autoref{fig:toi540_periodograms}, the SNR periodogram using \texttt{biweight+BLS} and \texttt{nuance} are very similar, with the known transiting exoplanet TOI-540 b detected with an orbital period P = 1.24 days. This is well expected as the relative parameter $\tau$ equal 13 for TOI-540 b \footnote{using \autoref{eq:relative_params} with the stellar rotation $P=0.72$ days and the known transit duration $D=29.5$ minutes of TOI-540 b \citep{TOI540}}, which lies outside the range where \nuancemethod{} is expected to be beneficial beneficial (see \autoref{fig:snr_detrend}). Nonetheless, the \nuance{} periodogram of TOI 540 features less spurious SNR peaks, largely due to the penalty naturally occurring when single transits with different depths are periodically combined. In the second search (right panel) of \autoref{fig:toi540_periodograms}, we also notice a higher number of peaks that would lead to false positive detections of transits in the \texttt{biweight+BLS} case. The proper treatment of correlated noise in \nuance{}, as observed in \autoref{real}, makes these peaks non-significant, avoiding a large number of false detections.\\\\

@@

\subsection{The GP kernel}
While not being discussed in our study, \textbf{the efficiency of \nuancemethod{} to detect transits in correlated noise is highly dependant on the design of its GP kernel}. In the ensemble comparison of \autoref{real}, the goal was to choose a kernel and an optimization strategy suited to most of the studied light curves, leading to few outliers in the results that were indicative of a badly designed and/or optimized kernel. An alternative, recommended for more realistic blind searches, is to perform model comparison on well selected kernels, and to adapt the optimization strategy to each dataset.\\\\
When using \nuancecode{} on TESS light curves for example, it must be noted that the observed light curve variability might originate from a contaminated photometric aperture, so that a physically-interpretable GP kernel representing a single star activity is not necessarily appropriate. On the other hand, a single squared exponential GP kernel might also be sufficient for some applications.\\\\
Something to emphasize is that \nuancemethod{} cannot be used to produce reliable detrended light curves, as an optimal GP is often flexible enough to partially model transits (see \autoref{fig:issue2}). 
In contrast, the idea behind \nuancemethod{} is rather to compute the likelihood of data against a model containing both transits and correlated noise, without ever trying to disentangle both signals. In practice, it means that it can be very hard to actually verify the presence of transits found by \nuancemethod{} visually, so that transits may be detected but remain hidden in correlated noise. This is particularly true for stars displaying very high-frequency photometric pulsations (see the example in \autoref{fig:exocomets} where the detrended light curves dismissively follow the base model).

\subsection{Prospects}
The present implementation of \nuancemethod{} has the potential to be extended to be used beyond the search of periodic box-shaped transits. Here are ideas of possible use and extensions, from the most straightforward to the most ambitious:
\begin{enumerate}
    \item In order to compare \nuancecode{} to BLS-based methods, we injected and retrieved only box-shaped transits. However, similarly to the Transit-Least-Square algorithm from \cite{tls}, \textbf{limb-darkened transits can serve as base model} in the linear search and are expected to improve the transit search in the same way TLS provided an improvement over BLS.
    \item The linear search of \nuancemethod{} is a single-event detection algorithm that can be used to search for single transit events, but also \textbf{detect transiting exocomets and flares}, by simply changing the base astrophysical model in the last colum of the design matrix $\bf{X}$. While not being tested and benchmarked, \nuancecode{} already integrates this feature. An example of the detection of known transiting exocomets in $\beta$ Pictoris TESS light curves is shown in \autoref{fig:exocomets}.
    \item During the periodic search, no prior about the transit duration related to the orbital period of the planet was used. This was done in order to allow the detection of grazing transiting exoplanets that would produce shorter timescale transits compared to what is expected from a circular orbit with a null impact parameter. However, adding such priors might produce less spurious periodogram peaks and be very beneficial for the automatic search of transits in large datasets. Another idea, similar to the one employed in \cite{foreman2016}, is to leverage \textbf{model comparison in order to reject transits that are better described by the GP model alone}. Both ideas come at no cost given our modeling approach.
    \item Finally, the formalism of \nuancemethod{} could be adapted and used to \textbf{search for exoplanets featuring transit time variations (TTV)}. Indeed, this application only requires a modification of the periodic search, as maximum likelihood peaks close to linearly predicted transit epochs may be considered. This could be done either with a special nearest-neighbor algorithm or with a convolution of the computed likelihood grid with a Gaussian kernel. To maximize efficiency and interpretability, we would recommend these approaches to be explored analytically, rather than using a data-driven treatment of the linear search products.
\end{enumerate}

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{exocomet.pdf}
        \caption{Demonstration of the search of transiting exocomets in the TESS light curves of the star $\beta$ Pictoris (2 minutes PDCSAP data from sector 33). This star is known to display rapid $\delta$ Scuti type photometric variations with a period of about 30 minutes \citep{Lecavelier2022}. The GP kernel and its hyperparameters are chosen and optimized as in \autoref{real}.Here, we simply used the linear search of \nuancecode{} with a different base model, one that mimics the shape of transiting exocomets (with 20 trial durations), to compute the SNR time-series of the signal over one sector. The maximum-SNR events are displayed in the bottom of the figure and match with the ones found by \cite{Lecavelier2022}.}
        \end{centering}
        \label{fig:exocomets} 
\end{figure}

\section{Conclusion}
This paper presents \nuancemethod{}, an algorithm designed to detect planetary transits in light curves featuring correlated noise in the form of instrumental signals and stellar variability. In this context, a conventional approach involves detrending a light curve before searching for transits using a Box-Least-Square algorithm. However, we show that the incomplete modeling of a light curve for its detrending, one that ignores the presence of transits, degrades their signal-to-noise-ratio down to the point of not being detectable. Adopting a widely used detrending stratgey, we explore the extent of this degradation on simulated light curves, and its dependance on the photometric stellar variability characteristics, showing the need for a full-fledged transit search method like \nuancemethod{}.\\\\
The effectiveness of \nuancemethod{} is tested using a synthetic dataset and further validated on real TESS light curves of 438 rapidly-rotating M dwarfs. These injection-recovery tests reveal that \nuancemethod{} consistently outperforms conventional transit search techniques, especially for transits with a duration exceeding one-fifth of the stellar variability timescale. In these cases, \nuancemethod{} not only lead to a higher number of true positive detections but also minimizes false positives, demonstrating its robustness and reliability.\\\\
We make \nuancemethod{} publicly available through the \nuancecode{} open-source Python package, developped with \textsf{JAX} to allow its  use on distributed computing environments and GPU devices. Overall, we acknowledge the limitations of \nuancemethod{} and its increased computational cost compared to more conventional techniques. Hence, \nuancemethod{} should be used as an alternative to more traditional techniques only in the presence of substantial correlated noise. As guidelines for choosing our method over other techniques are highly dependant on the type of detrending algorithm employed, we reserve this study to a future work.\\\\
Finally, we suggest future improvements and extensions of the algorithm, including its application for detecting single transits, exocomets, flares, and exoplanets featuring transit time variations (TTV), underscoring its versatility and potential for broader impact in astronomical research.
\\\\
\vfill{}
We would like to thank Julien De Wit, Prajwal Niraula and Jon Jenkins for meaningful discussions at the beginning of this project, Germain Garcia for his useful insights about numerical optimization, Michaël Gillon for his overall support, and the member of the Astronomical Data Group at the Center for Computational Astrophysics for many enriching discussions and feedback.
